signalHsmm - a novel semi-Markov model of eukaryotic signal peptides

### 2015-06-16 paper.tex
cvres figure done
minor changes to hsmm section

Let us first define Hidden Markov Model (HMM). Suppose you have a sequence of observations e.g. amino acids. What is of interest is
understanding underlying cause of their occurrence. 
HMM aims to answer that by assuming a specific yet flexible structure of the problem.
HMM consists of two stochastic processes. First is discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
These are "the cause" for the observations. At every step, hidden state might change according to transition matrix
$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application hidden states are signal peptide regions.
Second process $E_{t=1}^T$ is observation process, defined on the set of possible observations $\{O_1, \dots, O_m\}$. 
Observations are assumed to occur independently. Their distribution depends only on hidden state which emits it. 
Probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
In our case, observations are (degenerated) amino acids.
Main goal in signalHsmm is to find, for a given peptide, most probable regions boundaries. This is achieved with
Viterbi algorithm.
For a good reference on HMM see~\citep{1989rabinera}.

In regular HMM, hidden state duration -- number of observations emitted by the hidden state, has geometric distribution.
\citep{Durbin98biologicalsequence} shows how to extend it for some different distributions without significant increase in computational 
complexity. Similar ideas were used for signal peptide recognition, for example in~\citep{2004klla}. 
It is However still not flexible enough. Empirical regions length distributions (see figure~\ref{fig:reglen})
are difficult to capture in that way.

@@

The model we use is Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing 
any given hidden state length distribution. Scheme is shown in figure~\ref{fig:hsmm}.
Model is given by matrices $A$ and $B$ as before, as well as probabilities of duration length in hidden states.
$$\mathcal{P}(\text{duration in state} = d | \text{state is } S_i), \;\; i = 1, \dots, n, \;\; d = 1, \dots, D$$
$D$ is maximal allowed duration.
As our datasets are of reasonable size, $D$ is small -- around 30, computational effort is not much 
higher than in the regular HMM.

The gain is not only a better fit to the data but also, as 
model structure (see figure~\ref{fig:model_hsmm}) is straightforward, easier interpretation 
for researcher without mathematical background.

@@

Using the biogram software (BIOGRAM CITATION) we extracted n-grams from cleavage sites of signal peptides. 
XXX n-grams were further used in the analysis

@@

\begin{figure}[ht]\centering
\includegraphics[width=0.9\textwidth]{figures/cvres.png}
\caption{Results of cross-validation for 15 best performing amino acid encodings.}
\label{fig:cvres}
\end{figure}

>>>>

Let us first focus on Hidden Markov Model (HMM). 
Suppose you have a sequence of observations e.g. amino acids. What is of interest is understanding underlying cause of their occurrence. 
HMM aims to answer that by assuming a specific yet flexible structure of the problem.
HMM consists of two stochastic processes. First is discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
These are "the cause" for the observations. At every step, hidden state might change according to transition matrix
$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application hidden states are signal peptide regions.
Second process $E_{t=1}^T$ is observation process, defined on the set of possible observations $\{O_1, \dots, O_m\}$. 
Observations are assumed to occur independently. Their distribution depends only on hidden state which emits it. 
Probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
In our case, observations are (degenerated) amino acids.
Main goal in signalHsmm is to find, for a given peptide, most probable regions boundaries. This is achieved with
Viterbi algorithm.
For a good reference on HMM see~\citep{1989rabinera}.

In regular HMM, hidden state duration -- number of observations emitted by the hidden state, has geometric distribution.
\citep{Durbin98biologicalsequence} shows how to extend it for some different distributions without significant increase in computational 
complexity. Similar ideas were used for signal peptide recognition, for example in~\citep{2004klla}. 
It is, however, still not flexible enough. Empirical regional length distributions (see figure~\ref{fig:reglen})
are difficult to capture in that way.

@@

The model we use is Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing 
any given hidden state duration distribution. Scheme is shown in figure~\ref{fig:hsmm}.
In addition to matrices $A$ and $B$, model is given by probabilities of duration length in hidden states.
$$\mathcal{P}(\text{duration in state} = d | \text{state is } S_i), \;\; i = 1, \dots, n, \;\; d = 1, \dots, D$$
where $D$ is maximal allowed duration.
As our datasets are of reasonable size, $D$ is small -- around 30, computational effort is not much 
higher than with the regular HMM.

The advantage of HSMM model is not only a better performance but also, as 
model structure (see figure~\ref{fig:model_hsmm}) is straightforward, easier interpretation 
for researcher without mathematical background.

@@

Using the biogram software~\citep{biogramPackage} we extracted n-grams from cleavage sites of signal peptides. 
XXX n-grams were further used in the analysis

@@

\begin{figure}[ht]\centering
\includegraphics[width=0.9\textwidth]{figures/cvres.png}
\caption{Results of cross-validation. 1. An encoding providing the best sensitivity (AUC = 0.9337). 2. A group of encodings with be best sensitivity/specificity ratio (mean AUC = 0.9650). 3. An encoding providing the best specificity (AUC = 0.9686).}
\label{fig:cvres}
\end{figure}
